\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-05 k-planar.tex
added comments

We first formally define $k$-planar graphs.  An \defin{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a curve\footnote{A \defin{curve} in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \defin{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \defin{crossing} in an embedded graph $G$ is a triple $(p,vw,xy)$ with $p\in\R^2$, $vw,xy\in E(G)$ and such that $p\in (vw\cap xy)\setminus\{v,w,x,y\}$. An embedded graph $G$ is \defin{$k$-plane} if each edge of $G$ takes part in at most $k$ crossings.  A (not necessarily embedded) graph $G'$ is \defin{$k$-planar} if there exists a $k$-plane graph $G$ isomorphic to $G'$. Under these definitions, $0$-planar graphs are exactly planar graphs and $0$-plane graphs are exactly plane graphs.

@@

\note{DW}{We can motivate \cref{1-planar} since treewidth 3 is best possible, and the planarity of $H$ is essential in the proof of \cref{1PlanarQueue}. Futher motivation comes from the open problem at the end of the paper. Shall we argue this in the response? Or do we remove \cref{sec-1-planar}? We could keep the statement of \cref{1-planar} in the paper and refer to the arXiv paper for the proof. What do you think? All this depends on whether we can fix the errors below. I vote to fix the errors, add more motivation for \cref{1-planar}, keep the theorem and the proof. }

@@

\note{DW}{I suggest we say that it follows from Euler's formula that $G$ is finite. (To see this, say $G$ has $n$ vertices, $m$ non-crossed edges, and $k$ crossings. Let $G'$ be the plane graph obtained from $G$ by adding a dummy vertex at each crossing. So $G'$ is a plane multigraph in which every face is bounded by a 3-cycle. It follows from Euler's formula that $m+4k = |E(G')| = 3( |V(G')|-2) = 3 ( n+k-2)$, implying	$m+k = 3 (n-2)$ and $|E(G)| = m+2k \leq 6(n-2)$.) }

@@

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ and $G'$ be defined as above.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For each integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that
		\begin{compactenum}
			\item No edge of $F$ is crossed by any edge of $G$; and
			\item $V(F)$ can be partitioned into $P_1,\ldots,P_k$, for some $k\in\{1,2,3\}$ such that for each $i\in\{1,\ldots,k\}$,
			\begin{compactenum}
				\item $F[P_i]$ is a path; and
				\item $|V(P_i)\cap L_j| \le 15$ for all integers $j\ge 0$.
			\end{compactenum}
		\end{compactenum}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\mathcal{P}=\{S_x : x\in V(H)\}$ such that:
	\begin{compactenum}
		\item $H$ is planar;
		\item for all integers $j\ge 0$, and all $x\in V(H)$, $|S_x\cap L_j|\le 15$;
		\item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$; and
		\item $H$ has a tree-decomposition in which every bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
	\end{compactenum}
\end{lem}

@@

\note{DW}{We do not want to uncross.}

>>>>

We first formally define $k$-planar graphs.  An \defin{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a curve\footnote{A \defin{curve} in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \defin{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \defin{crossing} in an embedded graph $G$ is a triple $(p,vw,xy)$ with $p\in\R^2$, $vw,xy\in E(G)$ and such that $p\in (vw\cap xy)\setminus\{v,w,x,y\}$. An embedded graph $G$ is \defin{$k$-plane} if each edge of $G$ takes part in at most $k$ crossings. \note{DW}{We need at most two edges cross at a single point, so we can add a dummy vertex of degree 4, right?}  A (not necessarily embedded) graph $G'$ is \defin{$k$-planar} if there exists a $k$-plane graph $G$ isomorphic to $G'$. Under these definitions, $0$-planar graphs are exactly planar graphs and $0$-plane graphs are exactly plane graphs.

@@

\note{DW}{We can motivate \cref{1-planar} since treewidth 3 is best possible, and the planarity of $H$ is essential in the proof of \cref{1PlanarQueue}. Futher motivation comes from the open problem at the end of the paper. We need to argue this in the response and in the paper. The alternative is to remove \cref{sec-1-planar}? We could keep the statement of \cref{1-planar} in the paper and refer to the arXiv paper for the proof. What do you think? All this depends on whether we can fix the errors below. I vote to fix the errors, add more motivation for \cref{1-planar}, keep the theorem and the proof. }

@@

\note{DW}{I suggest we say that it follows from Euler's formula that $G$ is finite. (To see this, say $G$ has $n$ vertices, $m$ non-crossed edges, and $k$ crossings. Let $G'$ be the plane graph obtained from $G$ by adding a dummy vertex at each crossing. So $G'$ is a plane multigraph in which every face is bounded by a 3-cycle. It follows from Euler's formula that $m+4k = |E(G')| = 3( |V(G')|-2) = 3 ( n+k-2)$, implying	$m+k = 3 (n-2)$ and $|E(G)| = m+2k \leq 6(n-2)$, which can also be shown by deleting one edge from each crossing point. ) }

@@

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ and $G'$ be defined as above.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that
		\begin{compactenum}
			\item No edge of $F$ is crossed by any edge of $G$; and
			\item $V(F)$ can be partitioned into $P_1,\ldots,P_k$, for some $k\in\{1,2,3\}$ such that for each $i\in\{1,\ldots,k\}$,
			\begin{compactenum}
				\item $F[P_i]$ is a path; and
				\item $|V(P_i)\cap L_j| \le 15$ for every integer $j\ge 0$.
			\end{compactenum}
		\end{compactenum}
	\note{DW}{Should the vertices already grabbed by $P_i$ be mentioned in the setup? The new paths $Q_1,Q_2,Q_3$ need to avoid the vertices already grabbed by $P_1,P_2,P_3$. Basically, I am concerned about what happens when distinct vertical paths in $T$ want to grab the same vertex.}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\mathcal{P}=\{S_x : x\in V(H)\}$ such that:
	\begin{compactenum}
		\item $H$ is planar;
		\item for every integer $j\ge 0$ and every $x\in V(H)$, $|S_x\cap L_j|\le 15$;
		\item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$
		\note{DW}{Should this be $P_i\subseteq S_{x_i}$? (since the path grabs other vertices into its part)}; and
		\item $H$ has a tree-decomposition in which every bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
	\end{compactenum}
\end{lem}

@@

\note{DW}{But we do not want to uncross. Our claim that ``every pair of edges cross at most once'' is not true. I am convinced the result is still correct. Just need to think  of a better way to write it. Each crossing is in the vicinity of a vertex that has degree $\leq\Delta$. }
